---
title: "IADB FX Rates Diagnostics"
author: "Cedric Antunes / Evaluasi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
fontsize: 10pt
geometry: margin=0.65in
params:
  manual_dir: "data/manual"
  figures_dir: "figures/fx"
  fx_file: "IADB_fx_rates_daily.csv"
  diagnostics_file: "IADB_fx_rates_daily_diagnostics.csv"
  source_counts_file: "IADB_fx_rates_daily_source_counts.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 5.0,
  out.width = "\\linewidth",
  dpi = 300
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(janitor)
  library(readr)
  library(scales)
  library(knitr)
  library(here)
})

manual_dir <- here(params$manual_dir)
figures_dir <- here(params$figures_dir)

dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

fx_path <- file.path(manual_dir, params$fx_file)
diagnostics_path <- file.path(manual_dir, params$diagnostics_file)
source_counts_path <- file.path(manual_dir, params$source_counts_file)

stopifnot(file.exists(fx_path))

options(knitr.kable.NA = "")

print_compact_table <- function(data, caption = NULL, digits = 2, font_size = "\\scriptsize") {
  if (knitr::is_latex_output()) {
    cat(font_size, "\n")
  }

  table_out <- knitr::kable(
    data,
    digits = digits,
    caption = caption,
    booktabs = FALSE
  )

  cat(table_out, sep = "\n")

  if (knitr::is_latex_output()) {
    cat("\n\\normalsize\n")
  }
}
```

# Purpose

This document reports diagnostics for the daily exchange-rate table used in the IADB SAP cost analysis.

The FX table is expected to contain daily exchange rates in the following direction:

```text
fx_rate_local_per_usd = local currency units per 1 USD
```

For example, if `BRL = 5.20`, then one US dollar equals 5.20 Brazilian reais.

The output of this document is diagnostic only. It checks coverage, missingness, source composition, and the plausibility of exchange-rate trends before the FX table is used in the cost and time-cost analysis.

# Load FX data

```{r load-fx}
fx_daily <- read_csv(
  fx_path,
  show_col_types = FALSE
) |>
  clean_names() |>
  mutate(
    currency = str_to_upper(str_squish(currency)),
    fx_date = as.Date(fx_date),
    fx_rate_local_per_usd = as.numeric(fx_rate_local_per_usd),
    fx_source = as.character(fx_source)
  ) |>
  arrange(currency, fx_date)

fx_diagnostics_from_file <- if (file.exists(diagnostics_path)) {
  read_csv(diagnostics_path, show_col_types = FALSE) |>
    clean_names()
} else {
  NULL
}

fx_source_counts_from_file <- if (file.exists(source_counts_path)) {
  read_csv(source_counts_path, show_col_types = FALSE) |>
    clean_names()
} else {
  NULL
}

glimpse(fx_daily)
```

# Coverage diagnostics

```{r coverage-diagnostics, results='asis'}
fx_diagnostics <- fx_daily |>
  group_by(currency) |>
  summarise(
    first_date = min(fx_date, na.rm = TRUE),
    last_date = max(fx_date, na.rm = TRUE),
    n_days = n(),
    first_rate = first(fx_rate_local_per_usd),
    last_rate = last(fx_rate_local_per_usd),
    min_rate = suppressWarnings(min(fx_rate_local_per_usd, na.rm = TRUE)),
    max_rate = suppressWarnings(max(fx_rate_local_per_usd, na.rm = TRUE)),
    missing_rates = sum(is.na(fx_rate_local_per_usd)),
    n_sources = n_distinct(fx_source, na.rm = TRUE),
    sources = paste(sort(unique(na.omit(fx_source))), collapse = "; "),
    .groups = "drop"
  ) |>
  mutate(
    min_rate = ifelse(is.infinite(min_rate), NA_real_, min_rate),
    max_rate = ifelse(is.infinite(max_rate), NA_real_, max_rate),
    pct_change_window = 100 * (last_rate / first_rate - 1),
    sources_short = case_when(
      str_detect(sources, "USD_identity") ~ "USD identity",
      str_detect(sources, "FRED") ~ "FRED",
      str_detect(sources, "fawazahmed0") ~ "Public FX API",
      TRUE ~ str_trunc(sources, 32)
    )
  )

fx_diagnostics |>
  arrange(currency) |>
  transmute(
    Curr = currency,
    Start = first_date,
    End = last_date,
    Days = n_days,
    First = round(first_rate, 3),
    Last = round(last_rate, 3),
    Missing = missing_rates,
    Change = round(pct_change_window, 2),
    Source = sources_short
  ) |>
  print_compact_table(
    caption = "FX Coverage and Rate Diagnostics",
    digits = 3,
    font_size = "\\scriptsize"
  )
```

```{r safety-checks, results='asis'}
n_total_rows <- nrow(fx_daily)
n_currencies <- n_distinct(fx_daily$currency)
n_missing_rates <- sum(is.na(fx_daily$fx_rate_local_per_usd))
n_nonpositive_rates <- sum(fx_daily$fx_rate_local_per_usd <= 0, na.rm = TRUE)

safety_summary <- tibble(
  Item = c(
    "Total FX rows",
    "Currencies",
    "Missing rates",
    "Non-positive rates",
    "First FX date",
    "Last FX date"
  ),
  Value = c(
    as.character(n_total_rows),
    as.character(n_currencies),
    as.character(n_missing_rates),
    as.character(n_nonpositive_rates),
    as.character(min(fx_daily$fx_date, na.rm = TRUE)),
    as.character(max(fx_daily$fx_date, na.rm = TRUE))
  )
)

safety_summary |>
  print_compact_table(
    caption = "FX Safety Summary",
    digits = 2,
    font_size = "\\small"
  )

stopifnot(n_missing_rates == 0)
stopifnot(n_nonpositive_rates == 0)
```

# FX source diagnostics

```{r source-diagnostics, results='asis'}
fx_source_counts <- fx_daily |>
  count(currency, fx_source, name = "n_days") |>
  group_by(currency) |>
  mutate(
    pct_days = 100 * n_days / sum(n_days),
    fx_source_short = case_when(
      str_detect(fx_source, "USD_identity") ~ "USD identity",
      str_detect(fx_source, "FRED") ~ "FRED",
      str_detect(fx_source, "fawazahmed0") ~ "Public FX API",
      TRUE ~ str_trunc(fx_source, 32)
    )
  ) |>
  ungroup() |>
  arrange(currency, desc(n_days))

fx_source_counts |>
  transmute(
    Curr = currency,
    Source = fx_source_short,
    Days = n_days,
    Pct = round(pct_days, 1)
  ) |>
  print_compact_table(
    caption = "FX Source Coverage by Currency",
    digits = 1,
    font_size = "\\scriptsize"
  )
```

```{r source-plot, fig.width=7.2, fig.height=4.8, out.width="\\linewidth"}
p_source <- fx_source_counts |>
  ggplot(aes(
    x = reorder(currency, n_days),
    y = n_days,
    fill = fx_source_short
  )) +
  geom_col(width = 0.75) +
  coord_flip() +
  labs(
    title = "FX Source Coverage by Currency",
    x = NULL,
    y = "Number of currency-days",
    fill = "FX source",
    caption = "Note: Counts show the number of currency-days supplied by each FX source in the final daily FX table."
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    plot.caption = element_text(size = 7)
  )

p_source

ggsave(
  filename = file.path(figures_dir, "fx_source_coverage_by_currency.png"),
  plot = p_source,
  width = 10,
  height = 5,
  dpi = 300
)
```

# Daily exchange-rate levels

```{r rate-levels-plot, fig.width=7.2, fig.height=8.5, out.width="\\linewidth"}
p_levels <- fx_daily |>
  ggplot(aes(
    x = fx_date,
    y = fx_rate_local_per_usd,
    group = currency
  )) +
  geom_line(linewidth = 0.55) +
  facet_wrap(~ currency, scales = "free_y", ncol = 2) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
  labs(
    title = "Daily Exchange Rates to the US Dollar",
    subtitle = "Local currency units per 1 USD",
    x = NULL,
    y = "Local currency per USD",
    caption = "Note: Higher values indicate local-currency depreciation relative to the US dollar."
  ) +
  theme_bw(base_size = 9) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    plot.subtitle = element_text(hjust = 0.5, size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 7),
    axis.title = element_text(size = 8),
    strip.text = element_text(face = "bold", size = 8),
    plot.caption = element_text(size = 7)
  )

p_levels

ggsave(
  filename = file.path(figures_dir, "fx_daily_exchange_rate_levels.png"),
  plot = p_levels,
  width = 10,
  height = 8.5,
  dpi = 300
)
```

# Indexed exchange-rate movement

This plot indexes each currency to 100 on its first observed date. Values above 100 indicate depreciation against the US dollar relative to the beginning of the FX window; values below 100 indicate appreciation.

```{r indexed-rates, fig.width=7.2, fig.height=8.5, out.width="\\linewidth"}
fx_indexed <- fx_daily |>
  group_by(currency) |>
  arrange(fx_date, .by_group = TRUE) |>
  mutate(
    first_rate_currency = first(fx_rate_local_per_usd),
    fx_index_100 = 100 * fx_rate_local_per_usd / first_rate_currency
  ) |>
  ungroup()

p_index <- fx_indexed |>
  filter(currency != "USD") |>
  ggplot(aes(
    x = fx_date,
    y = fx_index_100,
    group = currency
  )) +
  geom_hline(yintercept = 100, linetype = "dashed", linewidth = 0.35) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ currency, scales = "free_y", ncol = 2) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
  labs(
    title = "Indexed Exchange-Rate Movement",
    subtitle = "First available date = 100",
    x = NULL,
    y = "Index",
    caption = "Note: Values above 100 indicate depreciation relative to the first date in the FX table."
  ) +
  theme_bw(base_size = 9) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    plot.subtitle = element_text(hjust = 0.5, size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 7),
    axis.title = element_text(size = 8),
    strip.text = element_text(face = "bold", size = 8),
    plot.caption = element_text(size = 7)
  )

p_index

ggsave(
  filename = file.path(figures_dir, "fx_indexed_exchange_rate_movement.png"),
  plot = p_index,
  width = 10,
  height = 8.5,
  dpi = 300
)
```

# Percent change over the FX window

```{r pct-change-plot, fig.width=7.2, fig.height=5.2, out.width="\\linewidth", results='asis'}
pct_change_table <- fx_indexed |>
  group_by(currency) |>
  summarise(
    first_date = first(fx_date),
    last_date = last(fx_date),
    first_rate = first(fx_rate_local_per_usd),
    last_rate = last(fx_rate_local_per_usd),
    pct_change = 100 * (last_rate / first_rate - 1),
    .groups = "drop"
  ) |>
  arrange(desc(pct_change))

pct_change_table |>
  transmute(
    Curr = currency,
    Start = first_date,
    End = last_date,
    First = round(first_rate, 3),
    Last = round(last_rate, 3),
    Change = round(pct_change, 2)
  ) |>
  print_compact_table(
    caption = "Percent Change in Local Currency per USD over the FX Window",
    digits = 2,
    font_size = "\\scriptsize"
  )

p_pct_change <- pct_change_table |>
  filter(currency != "USD") |>
  ggplot(aes(
    x = reorder(currency, pct_change),
    y = pct_change
  )) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
  geom_col(width = 0.75) +
  coord_flip() +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Exchange-Rate Change over the FX Window",
    subtitle = "Percent change in local currency per 1 USD",
    x = NULL,
    y = "Percent change",
    caption = "Note: Positive values indicate depreciation against the US dollar; negative values indicate appreciation."
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    plot.subtitle = element_text(hjust = 0.5, size = 9),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 8),
    plot.caption = element_text(size = 7)
  )

p_pct_change

ggsave(
  filename = file.path(figures_dir, "fx_percent_change_window.png"),
  plot = p_pct_change,
  width = 9,
  height = 5,
  dpi = 300
)
```

# Combined comparison plot

```{r combined-comparison, fig.width=7.2, fig.height=5.5, out.width="\\linewidth"}
p_combined <- fx_indexed |>
  filter(currency != "USD") |>
  ggplot(aes(
    x = fx_date,
    y = fx_index_100,
    group = currency,
    linetype = currency
  )) +
  geom_hline(yintercept = 100, linetype = "dashed", linewidth = 0.35) +
  geom_line(linewidth = 0.6) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
  labs(
    title = "Indexed FX Movement across IADB Countries",
    subtitle = "First available date = 100",
    x = NULL,
    y = "Exchange-rate index",
    linetype = "Currency",
    caption = "Note: Higher values indicate local-currency depreciation relative to the US dollar."
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    plot.subtitle = element_text(hjust = 0.5, size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 8),
    axis.title = element_text(size = 8),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.key.width = grid::unit(0.8, "cm"),
    plot.caption = element_text(size = 7)
  )

p_combined

ggsave(
  filename = file.path(figures_dir, "fx_indexed_combined_comparison.png"),
  plot = p_combined,
  width = 10,
  height = 6,
  dpi = 300
)
```

# Missingness heatmap

```{r missingness-heatmap, fig.width=7.2, fig.height=4.8, out.width="\\linewidth"}
fx_missing_grid <- fx_daily |>
  mutate(
    missing_rate = is.na(fx_rate_local_per_usd)
  )

p_missing <- fx_missing_grid |>
  ggplot(aes(
    x = fx_date,
    y = currency,
    fill = missing_rate
  )) +
  geom_tile() +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
  scale_fill_manual(
    values = c("FALSE" = "grey80", "TRUE" = "black"),
    labels = c("Available", "Missing")
  ) +
  labs(
    title = "FX Rate Availability by Currency-Date",
    x = NULL,
    y = NULL,
    fill = "Status",
    caption = "Note: The expected final FX table should have no missing currency-date cells."
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 8),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    plot.caption = element_text(size = 7)
  )

p_missing

ggsave(
  filename = file.path(figures_dir, "fx_missingness_heatmap.png"),
  plot = p_missing,
  width = 10,
  height = 5,
  dpi = 300
)
```

# Final validation

```{r final-validation, results='asis'}
validation_summary <- tibble(
  Check = c(
    "No missing FX rates",
    "No non-positive FX rates",
    "All currencies have same number of dates",
    "USD is always equal to 1"
  ),
  Passed = c(
    sum(is.na(fx_daily$fx_rate_local_per_usd)) == 0,
    sum(fx_daily$fx_rate_local_per_usd <= 0, na.rm = TRUE) == 0,
    fx_daily |> count(currency) |> pull(n) |> n_distinct() == 1,
    fx_daily |>
      filter(currency == "USD") |>
      summarise(ok = all(fx_rate_local_per_usd == 1, na.rm = TRUE)) |>
      pull(ok)
  )
)

validation_summary |>
  print_compact_table(
    caption = "Final FX Validation Checks",
    digits = 2,
    font_size = "\\small"
  )

if (!all(validation_summary$Passed)) {
  stop("At least one FX validation check failed. Review the diagnostics before running Script 06.")
}
```

# Notes for SAP cost analysis

The resulting FX file can be used by `06_cost_fx_and_time_outcomes.R`.

Recommended interpretation:

- `fx_rate_local_per_usd` is local currency per one US dollar.
- Local costs are converted to USD as:

```text
cost_usd = cost_local / fx_rate_local_per_usd
```

- Dollarized countries use `USD = 1`.
- The `fx_source` field should be retained for auditability.
- For client-facing reporting, public API fallback rates should be transparently documented and, where possible, spot-checked against official or client-approved sources.
